---
title: "Aviation Wildlife Strikes Analysis"
author: "Fabian Rüger"
format:
  html:
    embed-resources: true
    theme: cosmo
    toc: true
    number-sections: true
    code-fold: true
---

This report presents some **basic analyses** of the aviation wildlife dataset available at [Kaggle](https://www.kaggle.com/datasets/dianaddx/aircraft-wildlife-strikes-1990-2023/data). The dataset contains information about wildlife strikes with military, commercial or civil aircrafts from 1990 to 2023.

```{r}
#| label: load-packages
#| output: false
library(tidyverse)
```

```{r}
#| label: load-data
#| output: false

# load data
faa_wildlife <- read_csv(
  "../data/raw/faa_wildlife_strikes_1990_2023.csv",
  guess_max = Inf
)

# clean column names
faa_wildlife <- janitor::clean_names(faa_wildlife)
```

The dataset has `{r} ncol(faa_wildlife)` columns and `{r} nrow(faa_wildlife)` observations.

## Wildlife Strike Incidents Over Time
@fig-wildlife-strikes-by-year shows a positive trend in the number of wildlife strike incidents over time. The number of reported incidents has generally increased from around 2,500 incidents in 1990 to over 15,000 incidents in recent years. This increase could be attributed to various factors such as improved reporting mechanisms, increased air traffic, or changes in wildlife populations near airports.

```{r}
#| label: fig-wildlife-strikes-by-year
#| fig-cap: "Total Wildlife Strike Incidents per Year"
#| fig-width: 7
#| fig-height: 4
faa_wildlife |>
  group_by(incident_year) |>
  summarise(total_strikes = n(), .groups = "drop") |>
  ggplot(aes(x = incident_year, y = total_strikes)) +
  geom_line(color = "#00688B") +
  geom_point(color = "#00688B") +
  scale_x_continuous(breaks = seq(1990, 2023, by = 2)) +
  scale_y_continuous(
    breaks = seq(0, 17500, by = 2500),
    labels = scales::label_number(big.mark = ",")
  ) +
  labs(
    title = "Total Wildlife Strike Incidents per Year",
    x = "Year",
    y = "Number of Strikes"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

@fig-wildlife-strikes-by-month shows the average number of wildlife strike incidents by month. There is a clear seasonal pattern, with the highest number of incidents occurring during the spring and summer months (April to August). This trend may be related to increased wildlife activity during these months, as well as higher air traffic volumes.

```{r}
#| label: fig-wildlife-strikes-by-month
#| fig-cap: "Average Number of Wildlife Strike Incidents by Month"
#| fig-width: 7
#| fig-height: 4
#| warning: false
faa_wildlife |>
  group_by(incident_month, incident_year) |>
  summarise(n = n()) |>
  summarise(mean_strikes = mean(n)) |>
  ggplot(aes(x = incident_month, y = mean_strikes)) +
  geom_line(color = "#00688B") +
  geom_point(color = "#00688B") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Average Number of Wildlife Strike Incidents by Month",
    x = "Month",
    y = "Number of Strikes"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Taking a Closer Look at Wildlife Strikes
The wildlife species involved in strikes are diverse, but a few species account for a significant portion of the incidents. @fig-wildlife-strikes-by-species highlights the top species responsible for about 75% of wildlife strike incidents. Birds such as gulls, geese and larks are among the most frequently reported species involved in strikes, likely due to their prevalence near airports and flight paths.

```{r}
#| label: fig-wildlife-strikes-by-species
#| fig-cap: "Top Species Responsible for 75% of Wildlife Strike Incidents"
#| fig-width: 7
#| fig-height: 5
strikes_by_species_top <- faa_wildlife |>
  count(
    species,
    sort = TRUE
  ) |>
  mutate(
    cumsum = cumsum(n),
    p_tot = cumsum / sum(n)
  ) |>
  filter(p_tot <= .75)

faa_wildlife |>
  filter(species %in% strikes_by_species_top$species) |>
  ggplot(aes(x = fct_rev(fct_infreq(species)))) +
  geom_bar(fill = "#00688B") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Top Species Responsible for 75% of Wildlife Strike Incidents",
    x = "Species",
    y = "Number of Strikes"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```