---
title: "Aviation Wildlife Analysis"
author: "Fabian Rüger"
format:
  html:
    embed-resources: true
    code-fold: true
  pdf: default
---

This report presents some **basic analyses** of the aviation wildlife dataset available at [Kaggle](https://www.kaggle.com/datasets/dianaddx/aircraft-wildlife-strikes-1990-2023/data). The dataset contains information about wildlife strikes with military, commercial or civil aircrafts from 1990 to 2023.

```{r}
#| label: load-packages
#| output: false
library(tidyverse)
```

```{r}
#| label: load-data
#| output: false

# load data
faa_wildlife <- read_csv(
  "../data/raw/faa_wildlife_strikes_1990_2023.csv",
  guess_max = Inf
)

# clean column names
faa_wildlife <- janitor::clean_names(faa_wildlife)
```

The dataset has `{r} ncol(faa_wildlife)` columns and `{r} nrow(faa_wildlife)` observations.

## Wildlife Strike Incidents Over Time
@fig-wildlife-strikes-over-time shows a positive trend in the number of wildlife strike incidents over time. The number of reported incidents has generally increased from around 2,500 incidents in 1990 to over 15,000 incidents in recent years. This increase could be attributed to various factors such as improved reporting mechanisms, increased air traffic, or changes in wildlife populations near airports.


```{r}
#| label: fig-wildlife-strikes-over-time
#| fig-cap: "Total Wildlife Strike Incidents per Year"
#| fig-width: 7
#| fig-height: 4
faa_wildlife |>
  group_by(incident_year) |>
  summarise(total_strikes = n(), .groups = "drop") |>
  ggplot(aes(x = incident_year, y = total_strikes)) +
  geom_line(color = "#00688B") +
  geom_point(color = "#00688B") +
  scale_x_continuous(breaks = seq(1990, 2023, by = 2)) +
  scale_y_continuous(
    breaks = seq(0, 17500, by = 2500),
    labels = scales::label_number(big.mark = ",")
  ) +
  labs(
    title = "Total Wildlife Strike Incidents per Year",
    x = "Year",
    y = "Number of Strikes"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

@fig-wildlife-strikes-by-month shows the average number of wildlife strike incidents by month. There is a clear seasonal pattern, with the highest number of incidents occurring during the spring and summer months (April to August). This trend may be related to increased wildlife activity during these months, as well as higher air traffic volumes.

```{r}
#| label: fig-wildlife-strikes-by-month
#| fig-cap: "Average Number of Wildlife Strike Incidents by Month"
#| fig-width: 7
#| fig-height: 4
#| warning: false
faa_wildlife |>
  group_by(incident_month, incident_year) |>
  summarise(n = n()) |>
  summarise(mean_strikes = mean(n)) |>
  ggplot(aes(x = incident_month, y = mean_strikes)) +
  geom_line(color = "#00688B") +
  geom_point(color = "#00688B") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Average Number of Wildlife Strike Incidents by Month",
    x = "Month",
    y = "Number of Strikes"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```